---
title: "Importing data from YAML file to R"
author: "Taha Ahmed"
date: "`r Sys.Date()`"
bibliography: references.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Importing data from YAML file to R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r load-packages}
#| echo: FALSE
#| message: FALSE
#| include: FALSE
library(here)
library(yaml)
library(dplyr)
library(tibble)
library(tidyr)
library(ggplot2)
library(knitr)
```


```{r global_options}
#| echo: FALSE
#| message: FALSE
options(
   digits   = 7,
   width    = 84,
   stringsAsFactors = FALSE)
yaml_path <- here("inst/extdata/periodicdata.yml")
```


The aim here should be to read the YAML data into an R tibble `periodicdata`.


## Reading the YAML data

We can use the `yaml::read_yaml()` function to ingest the YAML file.

```{r read-yaml}
# imported YAML data (list of lists of lists)
# read_yaml() uses the YAML parser provided by libyaml, which conforms to the YAML 1.1 specification
pd_yaml <- yaml::read_yaml(
   file = yaml_path,
   # we want as.named.list = TRUE, otherwise we lose the property names
   as.named.list = TRUE, # TRUE is default
   merge.warning = FALSE, # FALSE is default
   # merge.precedence has no effect for this dataset, as far as I can tell
   merge.precedence = "order") # "order" is default, else "override"
```

For starters, let's list all elemental properties that make up the *periodic data*:
```{r properties-yaml}
#| results: markup
pd_yaml |> summary()
```

Now we need to do some wrangling to get this R list into a shape we can work with,
primarily in order to rectangularize it and make it into one or more tibbles.

We can use `tidyr::unnest_longer()` to convert a list of named 1-element lists to
a two column tibble.

```{r yaml-to-tibbles}
#| results: markup
# create a list of tibbles, where each tibble holds a property and the following fields
# (only "value" or those with length == 1 to ensure they can be recycled)
pdl_fields <- c("title", "title_symbol", "unit", "description", "footnote", "value")
# this list is also named with each property from the get-go
# https://stackoverflow.com/a/35035142/1198249
# Periodic Data List of tibbles
pdl <- setNames(vector("list", length = length(pd_yaml)), names(pd_yaml))
# The "note" field cannot be expected to have the same number of rows as "value",
# so to make my life easier we will import that into a separate variable
pdl_note <- setNames(vector("list", length = length(pd_yaml)), names(pd_yaml))
for (p in 1:length(pd_yaml)) {
   pdl[[p]] <-
      pd_yaml[[p]][names(pd_yaml[[p]]) %in% pdl_fields] %>%
      tibble::as_tibble() %>%
      # https://tidyr.tidyverse.org/reference/unnest_longer.html
      tidyr::unnest_longer(col = value, indices_to = "Symbol") %>%
      # note that this "value" column automatically gets an attribute "names"
      # which is redundant since we have the "symbol" column. Get rid of it.
      # To drop attributes from all columns at once: mutate(across(everything(), as.vector))
      mutate(value = as.vector(value))
   # now let's take care of the "note" field
   if (any(names(pd_yaml[[p]]) %in% c("note"))) {
      pdl_note[[p]] <-
         pd_yaml[[p]][names(pd_yaml[[p]]) %in% c("note")] %>%
         tibble::as_tibble() %>%
         tidyr::unnest_longer(col = note, indices_to = "Symbol")
      # now we can easily put "note" into place in each tibble
      # by joining by "symbol" we ensure that each note goes on the correct element's row
      pdl[[p]] <- left_join(pdl[[p]], pdl_note[[p]], by = "Symbol")
   }
}
# at this point we no longer need the pdl_note list - all its data is incorporated
# in the tibbles of pdl
# print three random tibbles from the list (poor man's quality control)
pdl[sample(1:length(pdl), 3)]
```

So now we have a list that contains a tibble for each property.
All the property tibbles use the elemental `Symbol` as the unique column.

The `reference` dict needs special consideration.
Because it can be a list of arbitrary length with each element being
a dict of multiple named elements (`url`, `doi`, `citation`, possibly others),
stuffing this jumble into a tibble column seems would likely render it data unusable anyway.

What if we instead put the `reference` data into an attribute of the tibble?

+ <https://stackoverflow.com/questions/48157225/how-to-add-metadata-to-a-tibble>

I considered using `comment` instead of `attribute`, because the latter requires us to explicitly
state the name of the attribute. And I only expect `reference` to be attached to the tibble
in this manner, so let's save on some typing for everyone involved?
I would have preferred to use `comment`, but unfortunalely `comment` only works
with *character vectors*, and so cannot support our lists.

```{r yaml-reference-to-attribute}
for (p in 1:length(pd_yaml)) {
   if (any(names(pd_yaml[[p]]) %in% c("reference"))) {
      attr(pdl[[p]], "reference") <-
         pd_yaml[[p]][names(pd_yaml[[p]]) %in% c("reference")]$reference
   }
}
```

Note that `attribute`s are not printed when printing the tibble - to see it
you have to explicitly call `attr()`, for example:
```{r}
#| results: markup
pdl$Production_2010 |> attr("reference")
```


## Joining `value` from all elemental properties into a wide-form tibble

What I am looking for is a tibble with 118 rows (column `symbol`) where each
column is a property (for the moment without considering units).
The name of the column should preferably be the name of the list element
in `pdl` (because in contrast to the `title` column this name can be trusted
to not contain spaces).

How will this tibble behave when we start introducing properties that have more
than one value per element? Well, then we can simply let those columns be nested,
and unnest on-demand if necessary. Feels like this approach should work.

I want to allow the YAML data to not necessarily include all known elements
for every property (some properties are only defined for some elements, anyway).

> Why not a long-form tibble?
> `bind_rows()` would have been a good choice, but our data has a quirk:
> the same column name (`value` in this case) can have different type across
> properties, being character in one and numeric in another.
> This severely hampers *any* attempt to construct a long-form data frame.


```{r wide-tibble-every-property}
#| echo: TRUE
# First, we need something to join against, for this I chose Symbol which is
# a natural choice as the unique identifier in this dataset
# For the sake of consistency, use capitalized name for Symbol column
pdt <- tibble(Symbol = pdl$Symbol$value)
for (p in which(names(pdl) != "Symbol")) {
   pdt <- left_join(
      x = pdt,
      y = pdl[[p]] |>
         # discard all columns except $value
         select(value, Symbol) |>
         # value in column $title should be set as new name for column $value.
         # Turns out it is doable, but note the unusual syntax !! :=
         # https://stackoverflow.com/questions/45472480/how-to-rename-a-column-to-a-variable-name-in-a-tidyverse-way
         dplyr::rename(!!names(pdl)[p] := value),
      by = "Symbol")
}
```

Print a random selection of columns (to aid in error-checking while rebuilding the vignette).
We randomly reshuffle the order of columns except for the first two, this way the info printed
at the bottom of the tibble output still state the correct number of non-visible columns.

+ <https://statisticsglobe.com/randomly-reorder-data-frame-by-row-and-column-in-r#example-2-shuffle-data-frame-by-column>

```{r}
#| results: markup
pdt[, c(1, 2, sample(3:ncol(pdt)))]
```



## The IUPAC periodic table

So let's demonstrate how to put together a periodic table from this data.
We will use ggplot2 for plotting, and first we will need to construct a new
tibble built containing the elemental properties we require, namely,
`Symbol`, `Name`, `Atomic_Number`, `Conventional_Atomic_Weight`, `Group`,
`Period`, and `Series`.

```{r periodictable-dimensions}
#| echo: false
# dimensions (sensitive setting, box alignment will break if changed)
ptable.width  <- 9.00
ptable.height <- 5.25
element.symbol.voffset <- 0.1
element.name.voffset   <- 0.12
element.period.voffset <- 0.32
# adjusted vertical position for lanthanide period
lanthanoids.vpos <- 8.5
# adjusted vertical position for actinide period
actinoids.vpos <- 9.5
# position and size of key (legend, manual)
ptable.legend <- tibble(
   Group = 4, Period = 2,
   Symbol = "Symbol", Atomic_Number = "atomic number", Name = "name", Label = "Key:",
   Conventional_Atomic_Weight = "conventional atomic wt")
```


```{r periodictable-tibble}
ptable_iupac <-
   pdt %>%
   select(Symbol, Name, Atomic_Number, Conventional_Atomic_Weight, Group, Period, Series) %>%
   # adjust vertical position for La and Ac periods *before* changing type to character
   # https://stackoverflow.com/questions/22337394/dplyr-mutate-with-conditional-values
   mutate(Period = case_when(
      # Atomic_Number >= 57 & Atomic_Number <= 71 ~ lanthanides.vpos,
      Series == "Lanthanide" ~ lanthanoids.vpos,
      # Atomic_Number >= 89 & Atomic_Number <= 103 ~ actinides.vpos,
      Series == "Actinide" ~ actinoids.vpos,
      TRUE ~ Period)) %>%
   mutate(Atomic_Number = as.character(Atomic_Number)) %>%
   # create two virtual elements as placeholders for the lanthanoids/lanthanoids labels
   tibble::add_row(
      Name = c("lanthanoids", "actinoids"),
      Group = c(3, 3),
      Period = c(6, 7),
      Atomic_Number = c("57-71", "89-103"),
      Series = c("Lanthanide", "Actinide"))
```


```{r periodictable-iupac-ggplot, fig.width=ptable.width, fig.height=ptable.height}
#| echo: FALSE
#| warning: FALSE
## Warning: Removed 2 rows containing missing values or values outside the scale range
## (`geom_text()`).
## Warning: Removed 34 rows containing missing values or values outside the scale range
## (`geom_text()`).
ggplot() +
   ### LEGEND
   # key legend (manual)
   # a rectangular shape would be better... # ggstar extension? or just define a path?
   geom_point(
      data = ptable.legend,
      size = 22, shape = 0,
      aes(y = Period, x = Group)) +
   # key label
   geom_text(data = ptable.legend,
      size = 2.0, colour = "black",
      # manually adjusted until it aligned nicely with outer top-left of legend box
      aes(label = Label, y = Period - 0.8, x = Group - 0.5)) +
   # key atomic number
   geom_text(data = ptable.legend,
      size = 1.8, colour = "black",
      aes(label = Atomic_Number, y = Period - 1.1 * element.period.voffset, x = Group)) +
   # key element symbol
   geom_text(data = ptable.legend,
      size = 3, colour = "black", fontface = "bold",
      aes(label = Symbol, y = Period - element.symbol.voffset, x = Group)) +
   # key element name
   geom_text(data = ptable.legend,
      size = 1.6, colour = "black",
      aes(label = tolower(Name), y = Period + element.name.voffset, x = Group)) +
   # key atomic weight
   geom_text(data = ptable.legend,
      size = 1.6, colour = "black",
      aes(label = Conventional_Atomic_Weight, y = Period + 1.1 * element.period.voffset, x = Group)) +
   ### COPYLEFT
   annotate("text",
      x = 0.35, y = 9.5, vjust = 0.5, hjust = 0, size = 7.5,
      label = "ðŸ„") +
   annotate("text",
      x = 0.85, y = 9.5, vjust = 0.5, hjust = 0, size = 2.6,
      label = paste0("Last updated ", Sys.Date(), "\nFreely licensed CC0")) +
   ### ELEMENTS
   scale_x_continuous(
      breaks = seq(
         min(ptable_iupac$Group),
         max(ptable_iupac$Group)),
      limits = c(
         min(ptable_iupac$Group) - 1.2,
         max(ptable_iupac$Group) + 0.8),
      expand = c(0, 0)) +
   scale_y_continuous(
      trans = "reverse",
      breaks = seq(
         min(ptable_iupac$Period),
         max(ptable_iupac$Period)),
      limits = c(
         max(ptable_iupac$Period) + 1,
         min(ptable_iupac$Period) - 1.5),
      expand = c(0, 0)) +
   # lanthanoids and actinoids background colour drawn before boxes
   geom_point(
      data = ptable_iupac %>% filter(Series == "Lanthanide"),
      size = 16,
      shape = 15,
      colour = "#cfc0c7",
      aes(y = Period, x = Group)) +
   geom_point(
      data = ptable_iupac %>% filter(Series == "Actinide"),
      size = 16,
      shape = 15,
      colour = "#a58394",
      aes(y = Period, x = Group)) +
   # boxes for all elements
   geom_point(
      data = ptable_iupac,
      # size 16 leaves no gaps between the element boxes at fig.width=9,fig.height=5.25
      size = 16,
      shape = 0,
      aes(y = Period, x = Group)) +
   # element symbol
   geom_text(
      data = ptable_iupac,
      size = 3,
      colour = "black",
      fontface = "bold",
      aes(label = Symbol, y = Period - element.symbol.voffset, x = Group)) +
   # atomic number
   geom_text(
      data = ptable_iupac,
      size = 1.8,
      colour = "black",
      aes(label = Atomic_Number, y = Period - element.period.voffset, x = Group)) +
   # element name
   geom_text(
      data = ptable_iupac,
      size = 1.6,
      colour = "black",
      aes(label = tolower(Name), y = Period + element.name.voffset, x = Group)) +
   # atomic weight
   geom_text(
      data = ptable_iupac,
      size = 1.6,
      colour = "black",
      aes(label = Conventional_Atomic_Weight, y = Period + element.period.voffset, x = Group)) +
   # period labels
   geom_text(
      data = ptable_iupac %>% filter(Period <= 7) %>% filter(Group == 1),
      size = 1.7, colour = "black",
      aes(label = Period, y = Period),
      x = 0.38) +
   # group labels # positions manually adjusted
   geom_text(
      data = data.frame(
         y = c(0.38, 1.38, rep(3.38, 10), rep(1.38, 5), 0.38),
         x = seq(1, 18)),
      size = 1.7,
      colour = "black",
      aes(label = x, x = x, y = y)) +
   # title # position manually adjusted
   annotate("text",
      x = 9, y = 0.6, vjust = 0,
      size = 4.5,
      fontface = "bold",
      label = "IUPAC Periodic Table of the Elements") +
   # subtitle # position manually adjusted
   annotate("text",
      x = 9, y = 0.74, vjust = 1,
      size = 2.6,
      label = "Created with the periodicdata package") +
   theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      plot.margin = unit(c(0, 0, -0.85, -0.85), "line"),
      axis.title = element_blank(),
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      # center (roughly) over transition metal block
      legend.position.inside = c(0.42, 0.91),
      legend.justification = c(0.5, 1),
      legend.direction = "horizontal",
      # make the legend colourbar a little longer
      legend.key.width = unit(2.5, "line"),
      legend.title = element_blank(),
      legend.background = element_rect(fill = "transparent"))
```



```{r ggsave-periodictable-iupac-ggplot, eval=FALSE, echo=FALSE, warning=FALSE}
### NOTE eval should not be flipped manually, use "rebuild_package.sh --refresh-demo" instead!
# ggsave() saves the last generated plot by default
# create.dir=TRUE creates the directory even if it does not already exist
# PNG file uses much wider font than expected, causes overflow in some places!
ggplot2::ggsave(filename = here::here("doc", "periodictable-iupac.png"),
   create.dir = TRUE, width = ptable.width, height = ptable.height)
# ggsave to PDF fails when running build_vignettes() if plot contains Unicode characters
# but running render() works, although the CC0 symbol is not properly printed in PDF
# https://git.solarchemist.se/taha/periodicdata/issues/18
# https://stackoverflow.com/questions/62837580/save-unicode-characters-to-pdf-in-r
# also, font is slightly wider than expected (but not as wide as PNG)
ggplot2::ggsave(filename = here::here("doc", "periodictable-iupac.pdf"),
   create.dir = TRUE, width = ptable.width, height = ptable.height)
ggplot2::ggsave(filename = here("doc", "periodictable-iupac.svg"),
   create.dir = TRUE, width = ptable.width, height = ptable.height)
```




## Dynamically create a list of properties for the README


I want to include a table showing the properties in the README, which means
we are limited to Markdown format.

```{r, echo=FALSE, message=FALSE, results="asis"}
source(here::here("vignettes/code/generate-markdown.R"), local = knitr::knit_global())
```



## Create package data

```{r save-package-data, eval=FALSE, echo=TRUE}
### NOTE eval should not be flipped manually, use "rebuild_package.sh --refresh-data" instead!
periodicdata <- pdt
usethis::use_data(periodicdata, internal = FALSE, overwrite = TRUE)
# I think YAML format is much easier to work with than CSV in this case
# but just in case others don't agree, let's also write a CSV for each dataframe
readr::write_csv(periodicdata, here("inst/extdata", "periodicdata.csv"))
# generate R/data.R
source(here::here("vignettes/code/generate-roxygen.R"), local = knitr::knit_global())
```



## Links and notes

+ <https://pypi.org/project/PyYAML>
+ <https://github.com/renkun-ken/rlist>
+ <https://renkun-ken.github.io/rlist-tutorial>
+ <https://www.monicathieu.com/posts/2020-04-08-tidy-multilevel>
+ <https://www.spsanderson.com/steveondata/posts/2024-10-29>
+ <https://henrywang.nl/transform-list-into-dataframe-with-tidyr-and-purrr>
+ <https://www.spsanderson.com/steveondata/posts/rtip-2023-01-25/index.html>
+ <https://forum.posit.co/t/simplifying-tidy-workflow-with-list-of-tibbles-and-purrr/19387>
+ <https://pratikunterwegs.github.io/tres-tidy-tutorial/working-with-lists-and-iteration.html>
+ <https://purrr.tidyverse.org/reference/map.html>
+ <https://tidyr.tidyverse.org/reference/hoist.html>
+ <https://tidyr.tidyverse.org/articles/rectangle.html>
